---
title: "Batch & group effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Batch & group effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5, fig.align = "center", dpi = 96
)
library(synthetica)
library(tidyverse)
library(knitr)
library(kableExtra)
library(patchwork)
```

Two post-processors stamp designed structure onto an already-clean synthetic
dataset (a `data.frame` or a `synthetica_sim`). They share the same additive,
SD-unit shift mechanism but model different things:

* `add_batch_effect()` -- a *technical / nuisance* batch factor (the "which
  plate was this run on" structure), for stress-testing batch-correction.
* `add_group_effect()` -- a *designed biological* grouping factor whose levels
  are genuinely **nominal** (each perturbs its own features in its own way).

We start from a baseline synthetic `iris`:

```{r base-sim}
syn <- simulate_dataset(iris, n = 300, seed = 42, verbose = FALSE)
```

# Technical batch effects

## Uniform shifts: the outlier-plate scenario

`shifts` as a numeric vector of length `n_batches` moves *every* target feature
in a batch by the same SD-multiple. Here three batches are indistinguishable and
the fourth is shifted hard -- the classic outlier plate:

```{r batch-uniform}
batched <- add_batch_effect(syn, n_batches = 4,
                            shifts = c(0, 0, 0, 3), seed = 42)
table(batched$data$batch)
```

```{r batch-uniform-pca, fig.width = 12, fig.height = 5}
## syn data PCA 
num_syn <- syn$data[, vapply(syn$data, is.numeric, logical(1))]
pca_syn <- prcomp(num_syn, scale. = TRUE)$x

df_syn <- data.frame(pca_syn, Species = syn$data$Species)

p1 = ggplot(df_syn, aes(x = PC1, y = PC2, color = Species)) +
  geom_point(alpha = 0.8) +
  labs(title = "Original Simulated Data") +
  theme_minimal()

### batched data PCA 
num_batched <- batched$data[, vapply(batched$data, is.numeric, logical(1))]
pca_batched <- prcomp(num_batched, scale. = TRUE)$x

df_batched <- data.frame(pca_batched, batch = batched$data$batch, Species = batched$data$Species)

p2 = ggplot(df_batched, aes(x = PC1, y = PC2, color = Species, shape = batch)) +
  geom_point(alpha = 0.8) +
  labs(title = "Uniform batch effect (batch4 = +3 SD)",
       x = "PC1", y = "PC2") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") # Clean, distinct colors for your 4 batches

p1 | p2
```


## Spike-and-slab: realistic heterogeneous shifts

Real batch effects do not hit every feature equally. `shifts = "spike_slab"`
draws a per-feature shift for each batch -- `hit ~ Bernoulli(shift_prob)` then
`N(0, shift_sd)` -- leaving a clean `reference_batch`. `proportions` gives
unequal batch sizes. This is the realistic 'omics signature and scales to wide
matrices.

```{r batch-spike-slab}
batched_ss <- add_batch_effect(
  syn, 
  n_batches = 4, 
  shifts = "spike_slab",
  shift_prob = 0.5, # per feature probability being shifted
  shift_sd = 1, # the SD shift
  proportions = c(0.4, 0.3, 0.2, 0.1), # relative size of each batch; sums to 1
  reference_batch = 1, 
  seed = 6
)

cat("Batch sizes\n")
table(batched_ss$data$batch)                       # unequal sizes
cat("\nSift Matrix: Batch by featuer in SD units of shift from reference\n\n")
round(batched_ss$batch_effect$shift_matrix, 2)     # ground-truth shifts (row 1 = clean)
```

The recorded `shift_matrix` is the ground truth: hand the batched data to a
correction method and check how well it recovers the unshifted values.

# Designed group effects (nominal)

`add_group_effect()` adds a named grouping factor whose levels move different
features in different directions -- a genuinely nominal factor that the
single-latent copula could not produce (categorical phenotype injection can only
do an *ordinal* gradient). Two ways to specify the effect.

## Explicit: hand-tuned, non-monotone

A named list keyed by level, each a named numeric of per-feature shifts (SD
units). Omitted level = reference; omitted feature = 0. Here clay has small
petals while silt has wide sepals -- two different features, no ordering:

```{r group-explicit}
g <- add_group_effect(
  syn, name = "soil", levels = c("clay", "sand", "silt"),
  probs = c(0.5, 0.3, 0.2), # relative size of each batch; sums to 1
  effects = list(clay = c(Petal.Length = -0.9),
                 silt = c(Sepal.Width  =  1.0)),
  seed = 3
)

cat("Median Petal Length by Soil type\n")
tapply(g$data$Petal.Length, g$data$soil, median)   # clay is low
cat("\nMedian Sepal Width by Soil type\n")
tapply(g$data$Sepal.Width,  g$data$soil, median)   # silt is high
```

## Generative: scales to 'omics widths

A named-list matrix is infeasible for an 11,000-feature panel. The generative
mode says "each level perturbs a *proportion* of features at some *effect-size
distribution*" -- spike-and-slab drawn independently per level, so each level
gets its own random signature:

```{r group-generative}
## Generate a fake omics data set
omics <- as.data.frame(matrix(rnorm(300 * 500), 300, 500,
                              dimnames = list(NULL, paste0("p", 1:500))))

## Define a new variable `tissue` and add group effects
gg <- add_group_effect(
  omics, name = "tissue", 
  levels = c("liver", "muscle", "fat"),
  probs = c(0.5, 0.3, 0.2), # relative size of each batch; sums to 1
  effect_prop = 0.10, ## probability that a feature is effected
  effect_sd = 0.5, 
  seed = 42
)

## Number of features effected by each level
cat("Number of features influenced by each feature:\n")
rowSums( attr(gg, "group_effect")$shift_matrix != 0)   # ~50 of 500 features per level
```

Because each level draws its own random subset, the perturbed feature *sets*
barely overlap between levels -- the defining property of a nominal factor.
Affected feature marginals become mixtures across groups, which is intended:
the group genuinely changes the distribution.
