---
title: "synthetica: a walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{synthetica: a walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  dpi = 96
)
```

```{r init, include = FALSE}
library(synthetica)
library(tidyverse)
library(knitr)
library(kableExtra)

# Try to fetch the Olink npx_data1.rda for the realistic 'omics section.
# Falls back gracefully if there's no internet at build time.
olink_url  <- "https://raw.githubusercontent.com/Olink-Proteomics/OlinkRPackage/main/OlinkAnalyze/data/npx_data1.rda"
olink_path <- file.path(tempdir(), "npx_data1.rda")
olink_ok   <- tryCatch({
  if (!file.exists(olink_path)) {
    utils::download.file(olink_url, olink_path, quiet = TRUE, mode = "wb")
  }
  load(olink_path)             # provides `npx_data1`
  exists("npx_data1")
}, error = function(e) FALSE)
```

# What the `synthetica` R package is for

`synthetica` generates **synthetic copies of tabular datasets** that preserve
the marginal distribution of every column and the joint correlation
structure across columns -- without retaining any row-level information from
the input. It supports mixed-type data (numeric, binary, categorical) and is
designed to feel like a small, self-contained utility you reach for when you
need:

* A **shareable** version of cohort data, scrubbed of any individual rows
* **Teaching/test datasets** that mimic the statistical properties of real
  data without exposing it
* **Power analyses** with a known ground-truth effect baked in
* **Stress tests for batch-correction** or normalisation pipelines

The package exports four functions:

* [`simulate_dataset()`][simulate_dataset] -- the main copula simulator, with
  optional phenotype injection and per-level fitting (`stratify_by`)
* [`add_batch_effect()`][add_batch_effect] -- a post-processor that introduces
  a synthetic *technical* batch structure
* [`add_group_effect()`][add_group_effect] -- a post-processor that adds a
  designed *nominal* grouping factor
* [`simulate_longitudinal()`][simulate_longitudinal] -- a wrapper for
  repeated-measures / panel data

This article is the **getting-started tour**: baseline simulation, recovery
checks, stratification, integer/discrete fidelity, and a real proteomics
example. The other articles cover the rest:

* **Injecting phenotypes** -- continuous, binary, and categorical
  traits correlated with chosen features
* **Batch & group effects** -- technical batches and designed biological groups
* **Longitudinal data** -- repeated measures via `simulate_longitudinal()`


# Simple Simulation: 

## Simulate a synthetic copy of the `iris` dataset

Generate a synthetic copy of 300 rows, doubling the size of the input. 
  - note that we can ask for any `n`

```{r mode-a-run}
sim <- simulate_dataset(iris, 
                        n = nrow(iris) * 2, 
                        seed = 42, 
                        verbose = FALSE)
```


### `sim` is a list of class `r class(sim)`. 

The synthetic `data.frame` lives in `sim$data`: with `r nrow(sim$data)` rows and `r ncol(sim$data)` columns, it has the same shape as the input but no row-level information is retained. 

```{r sim_class}
class(sim)
```

### There are various attributes in the `sim` object:

```{r}
names(sim)
```

### The `data` attribute is the synthetic dataset itself:

```{r sim_data}
dim(sim$data)
```

### The `audit` attribute summarises what was dropped or collapsed by the simulations guardrails:

```{r, sim_audit}
sim$audit
```


## How well does the simulated match the input?

### Side-by-side marginal recovery:

```{r mode-a-marginals}
input_num <- iris[, vapply(iris,    is.numeric, logical(1))]
sim_num   <- sim$data[, vapply(sim$data, is.numeric, logical(1))]

summary_compare <- data.frame(
  variable = names(input_num),
  in_median = vapply(input_num, median, numeric(1), na.rm = TRUE),
  sim_median= vapply(sim_num,   median, numeric(1), na.rm = TRUE),
  in_sd     = vapply(input_num, sd,     numeric(1), na.rm = TRUE),
  sim_sd    = vapply(sim_num,   sd,     numeric(1), na.rm = TRUE),
  in_miss   = vapply(input_num, function(x) mean(is.na(x)), numeric(1)),
  sim_miss  = vapply(sim_num,   function(x) mean(is.na(x)), numeric(1))
)

summary_compare |>
  kable(digits = 3,
        caption = "Marginal recovery: input vs simulated (numeric columns)") |>
  kable_styling(full_width = FALSE, font_size = 11)
```

### Categorical recovery (Species)

```{r mode-a-categorical}
# Categorical recovery (Species)
data.frame(
  level      = levels(iris$Species),
  input_prop = as.numeric(prop.table(table(iris$Species))),
  sim_prop   = as.numeric(prop.table(table(sim$data$Species)))
) |>
  kable(digits = 3, caption = "Species level frequencies") |>
  kable_styling(full_width = FALSE, font_size = 11)
```


### Joint correlation recovery

The Gaussian copula preserves rank (Spearman) correlations by construction:

```{r mode-a-correlation}
in_cor  <- cor(input_num, use = "pairwise.complete.obs", method = "spearman")
sim_cor <- cor(sim_num,   use = "pairwise.complete.obs", method = "spearman")

# Frobenius distance, normalised
fro <- norm(in_cor - sim_cor, type = "F") / norm(in_cor, type = "F")
cat(sprintf("Spearman correlation Frobenius distance (relative): %.4f\n", fro))
```


### Pairwise comparison for the strong pairs

```{r}
data.frame(
  pair       = c("Sepal.L vs Sepal.W", "Sepal.L vs Petal.L",
                 "Sepal.L vs Petal.W", "Sepal.W vs Petal.L",
                 "Sepal.W vs Petal.W", "Petal.L vs Petal.W"),
  input_cor  = c(in_cor[1,2],  in_cor[1,3],  in_cor[1,4],
                 in_cor[2,3],  in_cor[2,4],  in_cor[3,4]),
  sim_cor    = c(sim_cor[1,2], sim_cor[1,3], sim_cor[1,4],
                 sim_cor[2,3], sim_cor[2,4], sim_cor[3,4])
) |>
  kable(digits = 3, caption = "Pairwise Spearman correlations: input vs simulated") |>
  kable_styling(full_width = FALSE, font_size = 11)
```

The strong botanical correlations among iris features are reproduced very
closely.

## Stratified simulation

Iris is a textbook case where a single categorical column -- `Species` --
drives most of the variability in the continuous columns. A single Gaussian
copula fitted across all 150 rows captures the **global** marginal of each
variable and the linear correlation across them, but the three within-species
clusters blur together: the synthetic data won't quite reproduce the
characteristic three-cluster shape on PCA.

The fix is to **stratify** the simulation: pass `stratify_by = "Species"`
and the simulator fits a separate copula within each level, then
concatenates the per-stratum sims.

```{r stratify-run}
sim_strat <- simulate_dataset(iris, n = 300, seed = 42, verbose = FALSE,
                              stratify_by = "Species")

# Per-level row counts in the output
table(sim_strat$data$Species)

# Per-level marginals and correlation matrices are kept separately
names(sim_strat$marginals)
dim(sim_strat$corr$setosa)
```

Compare on PCA:

```{r stratify-pca, fig.width = 12, fig.height = 4.5}
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

plot_iris_pca <- function(df, title) {
  num <- df[, vapply(df, is.numeric, logical(1))]
  ok  <- stats::complete.cases(num)
  pca <- prcomp(num[ok, ], scale. = TRUE)
  cols <- c(setosa = "#E41A1C", versicolor = "#377EB8", virginica = "#4DAF4A")
  plot(pca$x[, 1:2], col = cols[as.character(df$Species[ok])],
       pch = 19, cex = 0.7,
       xlab = sprintf("PC1 (%.1f%%)", 100 * pca$sdev[1]^2 / sum(pca$sdev^2)),
       ylab = sprintf("PC2 (%.1f%%)", 100 * pca$sdev[2]^2 / sum(pca$sdev^2)),
       main = title)
  legend("topright", legend = names(cols), col = cols, pch = 19, bty = "n", cex = 0.8)
}

plot_iris_pca(iris,           "Real iris (n = 150)")
plot_iris_pca(sim$data,       "Single copula (n = 300)")
plot_iris_pca(sim_strat$data, "Stratified by Species (n = 300)")
```

The single-copula version produces three blurry overlapping clusters; the
stratified version reproduces the crisp three-cluster shape of the real
iris data. Per-stratum medians and proportions are preserved within
sampling noise:

```{r stratify-medians}
do.call(rbind, lapply(levels(iris$Species), function(sp) {
  data.frame(
    Species             = sp,
    in_Petal_L_median   = median(iris$Petal.Length[iris$Species == sp], na.rm = TRUE),
    sim_Petal_L_median  = median(sim_strat$data$Petal.Length[sim_strat$data$Species == sp], na.rm = TRUE),
    in_Petal_W_median   = median(iris$Petal.Width[iris$Species == sp], na.rm = TRUE),
    sim_Petal_W_median  = median(sim_strat$data$Petal.Width[sim_strat$data$Species == sp], na.rm = TRUE)
  )
})) |>
  kable(digits = 3,
        caption = "Per-species median recovery (stratified simulation)") |>
  kable_styling(full_width = FALSE, font_size = 11)
```

When to reach for `stratify_by`: a categorical column has a strong effect
on the continuous columns and you care about preserving the within-group
joint structure (not just the global marginal and global correlation).
Each stratum needs at least `min_strata_n` rows (default 30); smaller
strata are skipped with a warning.

# Integer and discrete fidelity

`synthetica` detects column types and preserves them on the round-trip.
Whole-number columns come back as **integers** (not interpolated floats), and
the correlations among binary / ordered-discrete columns are recovered rather
than attenuated. `mtcars` is a compact mixed-type example:

```{r integer-types}
sim_cars <- simulate_dataset(mtcars, n = nrow(mtcars), seed = 42,
                             verbose = FALSE)

data.frame(
  column     = names(mtcars),
  input_type = vapply(mtcars,        function(x) class(x)[1], character(1)),
  sim_type   = vapply(sim_cars$data, function(x) class(x)[1], character(1))
) |>
  kable(caption = "Storage class: input vs simulated (count-like columns stay integer)") |>
  kable_styling(full_width = FALSE, font_size = 11)
```

Count-like columns (`cyl`, `hp`, `gear`, `carb`, `vs`, `am`) are returned as
`integer`; the genuinely continuous measurements (`mpg`, `disp`, ...) stay
`double`. Under the hood, low-cardinality integers keep their **exact support**
and binary / ordered columns have their latent correlations recovered via
tetrachoric / biserial (binary) and polychoric / polyserial (ordinal)
corrections -- so a 0/1 or ordered-factor column's associations survive the
copula round-trip.

# A real 'omics example: Olink NPX data

```{r olink-skip, eval = !olink_ok, results = "asis", echo = FALSE}
cat("> _The Olink `npx_data1` dataset could not be fetched at build time
    (no internet?), so this section is skipped. See the package's GitHub
    pages site for the rendered version._\n")
```

The chunks below illustrate the same workflow on a real proteomics dataset
-- `npx_data1` from
[OlinkAnalyze](https://github.com/Olink-Proteomics/OlinkRPackage), which we
fetched at the top of the vignette.

```{r olink-shape, eval = olink_ok}
str(npx_data1, max.level = 1)
length(unique(npx_data1$SampleID))
length(unique(npx_data1$Assay))
```

The dataset is in long format -- one row per (sample, assay). Convert to wide
so each row is a sample and each column is an assay:

```{r olink-pivot, eval = olink_ok, warning = FALSE}
# npx_data1 is a tibble; coerce to a base data.frame so reshape(), rownames<-,
# and positional subsetting behave predictably.
long  <- as.data.frame(npx_data1)[, c("SampleID", "Assay", "NPX")]
wide  <- stats::reshape(long, idvar = "SampleID", timevar = "Assay",
                        direction = "wide")
# tidy up the column names ("NPX.<assay>" -> "<assay>")
names(wide)[-1] <- sub("^NPX\\.", "", names(wide)[-1])
rownames(wide)  <- wide$SampleID
wide$SampleID   <- NULL
dim(wide)
wide[1:4, 1:5] |>
  kable(digits = 3, caption = "First 4 samples x 5 assays of the wide NPX matrix") |>
  kable_styling(full_width = FALSE, font_size = 11)
```

Simulate a synthetic copy:

```{r olink-sim, eval = olink_ok}
sim_olink <- simulate_dataset(wide, n = nrow(wide), seed = 42, verbose = FALSE)
cat("Input dims :", nrow(wide), "x", ncol(wide), "\n",
    "Sim dims   :", nrow(sim_olink$data), "x", ncol(sim_olink$data), "\n",
    "Dropped by guardrails:", length(sim_olink$dropped_cols),
    "(", length(sim_olink$dropped_high_missing), "high-missing,",
    length(sim_olink$dropped_near_constant), "near-constant)\n")
```

Quick marginal-recovery check on a random subset of assays:

```{r olink-marginals, eval = olink_ok}
common  <- intersect(names(wide), names(sim_olink$data))
in_med  <- vapply(wide[, common, drop = FALSE], median, numeric(1), na.rm = TRUE)
sim_med <- vapply(sim_olink$data, median, numeric(1), na.rm = TRUE)
cat(sprintf("median  recovery (Pearson cor across %d assays): %.4f\n",
            length(common), cor(in_med, sim_med)))
```

Now apply the four-batch effect on top of the synthetic Olink data:

```{r olink-batched, eval = olink_ok}
sim_olink <- add_batch_effect(sim_olink, n_batches = 4,
                              shifts = c(0, 0.05, -0.05, 0.6),
                              seed = 1)

num    <- sim_olink$data[, vapply(sim_olink$data, is.numeric, logical(1))]
ok     <- stats::complete.cases(num)
pca    <- prcomp(num[ok, ], scale. = TRUE)
cols   <- c("steelblue", "darkorange", "forestgreen", "firebrick")
plot(pca$x[, 1:2],
     col  = cols[sim_olink$data$batch[ok]],
     pch  = 19, cex = 0.8,
     xlab = sprintf("PC1 (%.1f%%)", 100 * pca$sdev[1]^2 / sum(pca$sdev^2)),
     ylab = sprintf("PC2 (%.1f%%)", 100 * pca$sdev[2]^2 / sum(pca$sdev^2)),
     main = "Olink synthetic copy + simulated batch effect")
legend("topright", legend = levels(sim_olink$data$batch),
       col = cols, pch = 19, bty = "n")
```

This is the headline use case for the package on 'omics data: realistic
correlation structure, realistic missingness, plus a controllable, known
batch effect for stress-testing downstream methods. See the **Batch & group
effects** article for the full `add_batch_effect()` / `add_group_effect()` API
(spike-and-slab shifts, unequal batch sizes, and designed biological groups).

# Privacy guardrails

`simulate_dataset()` applies a small set of guardrails before fitting, so
that synthetic output of real participant data does not memorise individuals.
The defaults are tuned for cohort-sized data:

| Knob | Default | Effect |
|---|---|---|
| `min_cell_size`      | 10   | Categorical levels with fewer obs are collapsed into `"other"` |
| `max_missingness`    | 0.90 | Columns above this rate are dropped (their tails cannot be faithfully simulated) |
| `drop_near_constant` | TRUE | Columns with < 2 distinct non-NA values are dropped |
| `marginal_resolution`| 128  | Quantile-grid knots for storing each continuous marginal -- the marginal is stored as a grid, never as the raw vector |
| `noise_on_corr`      | 0    | SD of Gaussian noise added to off-diagonals of the latent correlation matrix; set > 0 for a DP-flavoured guarantee |

Most of these are passed as a named list to the `privacy` argument;
`marginal_resolution` is a top-level argument. For example, to loosen the
dependence structure with correlation noise:

```{r privacy-knobs}
sim_dp <- simulate_dataset(
  iris, n = 150, seed = 42, verbose = FALSE,
  marginal_resolution = 128,
  privacy = list(min_cell_size = 10, noise_on_corr = 0.10)
)
```

`sim$audit` prints a one-line summary of what was dropped or collapsed,
which you can capture in a manifest before sharing synthetic output.

# Limitations

* **Binary phenotypes** retain a single, predictable thresholding
  factor `phi(c) / sqrt(p(1-p))` on the round-trip; the generating latent is
  spliced into the copula so the designed correlation is otherwise captured
  exactly. See `inject_truth$<name>$expected_sim_cor`.
* **Quantitative phenotypes targeting many highly-correlated features**
  attenuate mildly due to rank-deficient inversion of the target
  correlation matrix.
* **Unordered categorical features with > 2 levels** introduce ordering-
  dependent latent correlations. Marginal level frequencies are preserved
  exactly; cross-correlations are approximate. (Ordered discrete columns are
  polychoric-corrected.)
* **No causal / DAG structure** -- only joint correlations are preserved.
* **High-dimensional longitudinal 'omics** (features x timepoints in the
  thousands) is out of scope for `simulate_longitudinal()`; see the
  **Longitudinal data** article for the tractable regime.
