---
title: "Longitudinal data (repeated measures)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal data (repeated measures)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4, fig.align = "center", dpi = 96
)
library(synthetica)
library(knitr)
library(kableExtra)
```

# Repeated measures via wide-format

A plain copula treats every row as independent, which is wrong for repeated
measures: a subject's value at one timepoint is correlated with its value at the
next. `simulate_longitudinal()` handles this by reshaping to **wide** format
(one row per subject, a column per `feature x time`), reusing
`simulate_dataset()`, and reshaping back. You just name the `id` and `time`
columns.

This turns within-subject temporal autocorrelation into ordinary between-column
correlation, so everything applies for free: per-timepoint marginals capture the
trajectory shape (e.g. growth), the missingness model reproduces dropout, and
time-invariant covariates ride along as single columns.

## ChickWeight

`datasets::ChickWeight` is 50 chicks weighed over 12 timepoints, with a
time-invariant `Diet`:

```{r chick-sim}
data(ChickWeight)
sim <- simulate_longitudinal(ChickWeight, id = "Chick", time = "Time",
                             seed = 1, verbose = FALSE)

class(sim)        # "synthetica_long"
sim$static        # "Diet" auto-detected (constant within Chick)
sim$features      # "weight" (time-varying)
head(sim$data)
```

### The growth curve is preserved

Each timepoint has its own marginal, so the mean trajectory emerges
automatically:

```{r chick-growth}
data.frame(
  Time = sort(unique(ChickWeight$Time)),
  real = round(tapply(ChickWeight$weight,    ChickWeight$Time, mean), 1),
  syn  = round(tapply(sim$data$weight, sim$data$Time, mean), 1)
) |>
  kable(caption = "Mean weight by timepoint: real vs synthetic") |>
  kable_styling(full_width = FALSE, font_size = 11)
```

### Trajectories look like growth, not noise

```{r chick-spaghetti, fig.width = 8, fig.height = 4}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
cols <- c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3")
spaghetti <- function(df, title) {
  plot(NA, xlim = range(df$Time), ylim = range(df$weight, na.rm = TRUE),
       xlab = "Time", ylab = "weight", main = title)
  for (ch in unique(df$Chick)) {
    d <- df[df$Chick == ch, ]
    lines(d$Time, d$weight, col = adjustcolor(cols[as.integer(d$Diet)[1]], 0.5))
  }
}
spaghetti(ChickWeight, "Real")
spaghetti(sim$data,    "Synthetic")
par(op)
```

### Within-subject correlation is preserved

The point of the wide-format trick is that a subject's weights across time stay
coupled. Compare the day-0 vs day-20 correlation, aligned by chick:

```{r chick-cor}
pair_cor <- function(df, t0, t1) {
  a <- df[df$Time == t0, c("Chick", "weight")]
  b <- df[df$Time == t1, c("Chick", "weight")]
  m <- merge(a, b, by = "Chick")
  cor(m[[2]], m[[3]], use = "complete.obs")
}
c(real = round(pair_cor(ChickWeight, 0, 20), 3),
  syn  = round(pair_cor(sim$data,    0, 20), 3))
```

Chicks that died early reproduce as realistic **dropout** -- `(subject, time)`
rows with no observed feature are absent from the synthetic long output, just as
in the input.

## Other designs

The same recipe covers:

* **Before/after RCTs** -- two timepoints become `baseline` / `followup`
  columns; the pre-post correlation and any treatment shift are preserved.
* **Crossover trials** -- one column per period.
* **Modest biomarker panels** -- a handful of analytes over several visits.

## Scope

The wide frame has `features x timepoints` columns, so this approach is for the
tractable regime (up to a few hundred such columns); a warning fires beyond
that. High-dimensional longitudinal 'omics (thousands of features x several
timepoints) needs a different method -- subject-level random effects with an
autoregressive temporal residual structure -- and is out of scope here. The time
grid is the set of unique observed time values, so genuinely irregular or
continuous times should be binned to a common grid before calling.
