---
title: "Injecting phenotypes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Injecting phenotypes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5, fig.align = "center", dpi = 96
)
library(synthetica)
library(tidyverse)
library(knitr)
library(kableExtra)
library(patchwork)
```

# Inject a synthetic phenotype

Beyond reproducing the input, `simulate_dataset()` can **append new
synthetic columns** -- *phenotypes* -- each correlated with a chosen subset of
features at a user-defined effect size. This is useful for generating teaching
datasets, power analyses, or method benchmarks with a known ground-truth signal.

Each entry of `inject` is a named list describing the trait. The `effect`
controls how strongly it correlates with its `on` targets, either as a
standardized `mean_shift` or a target `r_squared` on the latent scale. Three
phenotype `type`s are supported: `"quantitative"`, `"binary"`, and
`"categorical"`.

```{r inject-all}
sim <- simulate_dataset(
  iris, n = 300, seed = 42, verbose = FALSE,
  inject = list(
    # 1. continuous trait, modest effect on every feature
    rainfall = list(
      type   = "quantitative", mean = 40, sd = 8,
      effect = list(kind = "r_squared", min_r2 = 0.35, on = "all")
    ),
    # 2. binary trait (case/control), stronger effect on the petals
    treated = list(
      type   = "binary", prob = 0.40,
      effect = list(kind = "r_squared", min_r2 = 0.3,
                    on = c("Petal.Length", "Petal.Width"))
    ),
    # 3. ordinal categorical trait: clay -> sand -> silt gradient
    soil = list(
      type   = "categorical",
      levels = c("clay", "sand", "silt"), probs = c(0.5, 0.3, 0.2),
      effect = list(kind = "r_squared", min_r2 = 0.3,
                    on = c("Petal.Length", "Petal.Width"))
    )
  )
)

head(sim$data, 6) |>
  kable(digits = 2, caption = "iris + three injected phenotypes") |>
  kable_styling(full_width = FALSE, font_size = 11)
```

## Variance Explained by Simulated Phenotype

```{r, fig.width = 6, fig.height = 4}
fit = lm(Petal.Length ~ rainfall + soil + treated, data = sim$data)
a = anova(fit)
eta_squared = a[, 2] / sum(a[,2])
names(eta_squared) = rownames(a)
barplot(eta_squared[-4], col = c("blue4","green4","brown"), 
        main = "Variance Explained by Simulated Phenotype", 
        ylab = "variance explained")

```


## Rainfall Correlations

```{r fig.width = 10, fig.height = 8}
p1 = sim$data |> ggplot(aes(x = rainfall, y = Petal.Length)) +
  geom_point( aes(color = Species) ) +
  # geom_smooth(method = "lm", aes(color = Species)) +
  geom_smooth(method = "lm", formula = 'y ~ x', 
              color = "grey", se = FALSE, 
              linetype = "dashed") +
  theme_bw()

p2 = sim$data |> ggplot(aes(x = soil, y = Petal.Length, fill = soil)) +
  geom_boxplot() +
  scale_fill_brewer(palette  = "Set1") +
  theme_bw()

p3 = sim$data |> 
  mutate(treated = as.factor(treated)) |>
  ggplot(aes(x = treated, y = Petal.Length, fill = treated)) +
  geom_boxplot() +
  scale_fill_brewer(palette  = "Dark2") +
  theme_bw()

p1 / (p2 | p3) + plot_layout(heights = c(2,1))
```


## What was realised vs requested

Every injected trait is reproducible bookkeeping in `sim$inject_truth`:

```{r inject-truth}
data.frame(
  phenotype        = names(sim$inject_truth),
  realised_cor     = vapply(sim$inject_truth,
                            function(it) it$mean_abs_real_cor, numeric(1)),
  expected_sim_cor = vapply(sim$inject_truth,
                            function(it) it$expected_sim_cor, numeric(1)),
  thresholding     = vapply(sim$inject_truth,
                            function(it) it$binary_sim_factor, numeric(1))
) |>
  kable(digits = 3, caption = "Injected phenotype: realised vs expected effect") |>
  kable_styling(full_width = FALSE, font_size = 11)
```

## Binary phenotypes: thresholded-normal fidelity

A binary trait is a continuous latent thresholded to 0/1. `synthetica` splices
that **generating latent** into the copula, so the designed correlation is
captured exactly -- only a single, predictable thresholding factor
`phi(c) / sqrt(p(1-p))` remains. `expected_sim_cor` (the latent target times
that factor) closely predicts the realised point-biserial in the output.

```{r inject-binary-check}
c(prevalence = mean(sim$data$treated == 1),
  is_integer = is.integer(sim$data$treated))
```

## Categorical phenotypes: an ordinal gradient

A `"categorical"` phenotype is an **ordinal** trait: one latent is correlated
with the targets and cut at the cumulative-probability thresholds of `probs`
into ordered `levels` (the first level is the low end). Because the relationship
is ordinal, the feature medians climb monotonically along the levels:

```{r inject-categorical-check}
prop.table(table(sim$data$soil))                       # ~ .5 / .3 / .2
tapply(sim$data$Petal.Length, sim$data$soil, median)   # clay < sand < silt
```

For a *nominal* factor -- where each level perturbs different features in
different directions, with no ordering -- use `add_group_effect()` (see the
**Batch & group effects** article). That is applied post-hoc, because a nominal
factor cannot survive the single-latent copula round-trip.

## Feasibility

Effects are checked for feasibility (`rho' R_X^-1 rho <= 1`). An infeasible
request -- e.g. a large effect on many mutually-correlated targets -- is scaled
down to the largest achievable magnitude, with a warning naming requested vs
achieved.
